---
title: "Climate plots - Demo figures"
format:
    html:
        embed-resources: true
        lightbox: true
        code-fold: true
---

```{r setup}
#| include: false
suppressPackageStartupMessages(library("tinyplot"))
suppressPackageStartupMessages(library("stars"))
suppressPackageStartupMessages(library("sf"))
suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("rnaturalearth"))
suppressPackageStartupMessages(library("colorspace"))
```

## Line Graph: Global Surface Temperature Change

This data set was used to create [Figure 3.3](https://doi.org/10.7927/606k-d497)
in the IPCC Assessment Report 6 (AR6) Synthesis Report.


```{r ipcc_load_data}
d <- readRDS("data/IPCC_AR6_global_surface_temp_change.rds")
title <- "Global surface temperature change relative to 1850-1900"
```

```{r ipcc_lineplot_1}
#| fig-width: 8
#| fig-height: 4
cols <- palette.colors(6, "R4")[-1]
tinyplot(Mean ~ Year | Scenario, type = "l", lwd = 2,
         col = cols,
         main = title,
         xlim = range(d$Year), ylim = range(d$Mean),
         data = subset(d, Scenario != "historical"))
tinyplot(Mean ~ Year, type = "l", lwd = 2, col = "black",
         data = subset(d, Scenario == "historical"),
         add = TRUE)
```

```{r ipcc_lineplot_2}
#| fig-width: 8
#| fig-height: 4
par(mar = c(5.1, 4.1, 4.1, 6.1), bty = "l")
#cols <- palette.colors(6, "R4")[-1]
cols <- hcl.colors(5, "Set 3")
tinyplot(Mean ~ Year | Scenario, type = "l", lwd = 2,
         col = cols,
         main = title,
         legend = FALSE,
         xlim = range(d$Year), ylim = range(d$Mean),
         data = subset(d, Scenario != "historical"))
tinyplot(Mean ~ Year, type = "l", lwd = 2, col = "black",
         data = subset(d, Scenario == "historical"),
         add = TRUE)
tmp <- subset(d, Year == max(Year))
for (i in seq_len(nrow(tmp))) {
    axis(side = 4, at = tmp$Mean[i], font = 2,
         labels = bquote(.(tmp$Scenario[1])~degree*C),
         col.axis = cols[i], las = 1, lwd = NA)
}
```


# World bank

The world bank also provides gridde data set .. a lot.
This is just an example of the Standardized Precipitation Evaporation Index (SPEI),
a drought indicator.


```{r}
#| fig-width: 8
#| fig-height: 6
ne <- ne_countries(returnclass = "sf")

tas <- read_stars(file.path("data",
        "anomaly-tas-annual-mean_cmip6-x0.25_ensemble-all-ssp245_climatology_median_2040-2059.nc"))
lim <- range(as.data.frame(tas)[, 4])
ggplot() + geom_stars(data = tas) + theme_minimal() +
    scale_fill_continuous_sequential("ag_Sunset", p1 = 0.6, limits = lim, name = bquote(degree*C)) +
    geom_sf(data = ne, fill = NA, lwd = 0.6) +
    labs(title = "Anomaly of average mean temperature, 2040-2059") +
    labs(subtitle = "Reference period 1995-2014") +
    theme_minimal()
```


```{r}
#| fig-width: 8
#| fig-height: 6
ne <- ne_countries(returnclass = "sf")

prp <- read_stars(file.path("data",
        "anomaly-prpercnt-annual-mean_cmip6-x0.25_ensemble-all-ssp245_climatology_median_2040-2059.nc"))
lim <- quantile(as.data.frame(prp)[, 4], probs = c(0.001, 0.999), na.rm = TRUE)
lim <- c(-1, 1) * max(abs(lim))
#lim <- range(as.data.frame(prp)[, 4])
ggplot() + geom_stars(data = prp) + theme_minimal() +
    scale_fill_continuous_diverging("Blue-Red", rev = TRUE, p1 = 0.5, limits = lim, name = bquote("%")) +
    geom_sf(data = ne, fill = NA, lwd = 0.6) +
    labs(title = "Anomaly of relative precipitation change, 2040-2059") +
    labs(subtitle = "Reference period 1995-2014") +
    theme_minimal()
```

# UWZ Warnings

Had a quick look at the website - they provide their warnings as geojson.
Downloaded todays warnings (heavy snowfall, March 26 2026) and added the
geojson to the `data` folder. For more details see `UWZ` folder content.


```{r UWZ_loading_data}
#| fig-width: 8
#| fig-height: 6
AT   <- read_sf("data/UWZ_AT.geojson")
WARN <- read_sf("data/UWZ_warnings_2026-03-26.geojson")

# Original colors
cols <- c("#00ff39", "#ffff40", "#ffb430", "#ff001b", "#ff0cfa") |> setNames((-1):3)
labs <- rep("", length(cols)) |> setNames(names(cols))
ggplot() +
    geom_sf(aes(fill = as.factor(-1)), data = AT, col = NA) +
    geom_sf(aes(fill = as.factor(level)), data = WARN, col = NA) +
    geom_sf(data = AT, fill = NA, lwd = .2) +
    scale_fill_manual(values = cols, name = "", labels = labs) +
    theme_minimal() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot() +
    geom_sf(aes(fill = as.factor(-1)), data = AT, col = NA) +
    geom_sf(aes(fill = as.factor(level)), data = WARN, col = NA) +
    geom_sf(data = AT, fill = NA, lwd = .2) +
    scale_fill_manual(values = deutan(cols), labels = labs) +
    theme_minimal() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```

```{r}
#| fig-width: 8
#| fig-height: 6
ggplot() +
    geom_sf(aes(fill = as.factor(-1)), data = AT, col = NA) +
    geom_sf(aes(fill = as.factor(level)), data = WARN, col = NA) +
    geom_sf(data = AT, fill = NA, lwd = .2) +
    scale_fill_manual(values = desaturate(cols), labels = labs) +
    theme_minimal() +
    theme(legend.position = "bottom", legend.direction = "horizontal")
```




