---
title: Base R Shows Its Colors
author:
  # see ?rjournal_article for more information
  - name: Achim Zeileis
    affiliation: Universität Innsbruck
    address:
    - Department of Statistics
    url: https://www.zeileis.org/
    orcid: 0000-0003-0918-3766
    email:  Achim.Zeileis@R-project.org
  - name: Paul Murrell
    affiliation: University of Auckland
    address:
    - Department of Statistics
    url: https://www.stat.auckland.ac.nz/~paul/
    orcid: 0000-0002-3224-8858
    email:  paul@stat.auckland.ac.nz
abstract: >
  Starting from R 4.0.0 R has a new and much improved default palette.
  More than 100 well-established palettes provided in `palette.colors()` and `hcl.colors()`.
  Overview and guidance.
preamble: |
  \let\proglang=\text
  
# per R journal requirement, the bib filename should be the same as the output 
# tex file. Don't forget to rename the bib file and change this example value.
bibliography: color.bib
output: rticles::rjournal_article
---

```{r preliminaries, include = FALSE}
library("colorspace")
```

# Color palettes in (base) R

## History

Color is an integral part of data visualization and statistical graphics.

Although graphics were emphasized to be one of the strengths of R already early on [@color:Ihaka+Gentleman:1996],
base R provided shockingly few functions for choosing colors from readily available palettes. And those palettes
that were provided (see Figure \ref{fig:oldbase}) were standard when they were implemented but turned out to be
rather poor.

- The colors from the default `palette()` could be accessed through numbers `1` to `8` and were mostly used for coding _categorical_ information.
- The functions `heat.colors()`, `terrain.colors()`, `topo.colors()`, and `gray.colors()` were mostly used as _sequential_ palettes for ordered or numeric information.
- The function `cm.colors()` was the only _diverging_ palette.
- The function `rainbow()` implements the infamous rainbow (or jet) color palette that was very widely used for all types, possibly after restricting the hue range: _categorical_, _sequential_, and _diverging_.

```{r oldbase, echo=FALSE, fig.height=1.75, fig.width=13, out.width='\\textwidth', fig.align='center', fig.pos='htbp', fig.cap='Old base R palettes.'}
oldbase.pals <- function() {
  c("default (R3)", "rainbow", paste0(c("heat", "terrain", "topo", "cm", "gray"), ".colors"))
}

oldbase.colors <- function(n, palette = "default", ...) {
  palette <- match.arg(palette, oldbase.pals())
  if(palette == oldbase.pals()[1]) {
    palette.colors(n, palette = "R3", ...)
  } else {
    do.call(palette, list(n = n, ...))
  }
}

oldbase.swatch <- function(palette = oldbase.pals(), n = 8, nrow = 2,
                           border = "black", cex = 1, ...)
{
     cols <- sapply(palette, oldbase.colors, n = n)
     ncol <- ncol(cols)
     nswatch <- min(ncol, nrow)
     op <- par(mar = rep(0.1, 4),
               mfrow = c(1, min(5, ceiling(ncol/nrow))),
     	       cex = cex, ...)
     on.exit(par(op))
     while (length(palette)) {
 	subset <- seq_len(min(nrow, ncol(cols)))
 	plot.new()
 	plot.window(c(0, n), c(0.25, nrow + 0.25))
 	y <- rev(subset)
 	text(0, y + 0.1, palette[subset], adj = c(0, 0))
 	y <- rep(y, each = n)
 	rect(rep(0:(n-1), n), y, rep(1:n, n), y - 0.5,
 	     col = cols[, subset], border = border)
 	palette <- palette[-subset]
 	cols    <- cols [, -subset, drop = FALSE]
     }
}

oldbase.swatch()
```

However, all of these palettes -- except `gray.colors()` -- have significant problems: They are ugly and have poor perceptual properties.
The colors are highly saturated (garish, flashy) and unbalanced with respect to chroma and luminance. Most of them also use sequences
of hues obtained in the RGB (red-green-blue) space, leading to clustering of colors at the red, green, and blue primaries.

Better color palettes had been neglected in base R for a long time. There are various reasons for this:

- Functions like `colorRampPalette()` allowed users to create new palettes on the fly -- provided they had useful colors available upon which
  the palette could be based.
- Better palettes were easily available starting with \CRANpkg{RColorBrewer} [@color:RColorBrewer], later
  \CRANpkg{colorspace} [@color:Ihaka:2003; @color:Zeileis+Hornik+Murrell:2009], and somewhat more recently
  \CRANpkg{viridis} [@color:viridis], \CRANpkg{rcartocolor} [@color:rcartocolor], or \CRANpkg{scico} [@color:scico].
- Instead of using base graphics, frameworks like \CRANpkg{ggplot2} [@color:ggplot2] or \CRANpkg{lattice} [@color:lattice]
  that bring along their own color themes were adopted very widely.


## New base R palettes

`palette.colors()` for qualitative palettes [@color:grDevices2]

`hcl.colors()` for qualitative, sequential, and diverging palettes obtained in HCL (hue-chroma-luminance) space [@color:Wiki+HCL]
see @color:grDevices and @color:Zeileis+Fisher+Hornik:2020.

Both accompanied by a function to query the available palette names (`palette.pals()` and `hcl.pals()`, respectively).
Color swatches shown in Figures \ref{fig:palette}, \ref{fig:hcls}, and \ref{fig:hcld} which are also shown by running
the `example()` of the two functions which make basic `palette.swatch()` and `hcl.swatch()` functions available.
Here, focus on `palette.colors()`, more details on `hcl.colors()` later.

```{r palette, echo=FALSE, fig.height=3.5, fig.width=13, out.width='\\textwidth', fig.align='center', fig.pos='htbp', fig.cap='Qualitative palettes from \\code{palette.colors()}.'}
palette.swatch <- function(palette = palette.pals(), n = 8, nrow = 4,
                           border = "black", cex = 1, ...)
{
     cols <- sapply(palette, palette.colors, n = n)
     ncol <- ncol(cols)
     nswatch <- min(ncol, nrow)
     op <- par(mar = rep(0.1, 4),
               mfrow = c(1, min(5, ceiling(ncol/nrow))),
     	       cex = cex, ...)
     on.exit(par(op))
     while (length(palette)) {
 	subset <- seq_len(min(nrow, ncol(cols)))
 	plot.new()
 	plot.window(c(0, n), c(0.25, nrow + 0.25))
 	y <- rev(subset)
 	text(0, y + 0.1, palette[subset], adj = c(0, 0))
 	y <- rep(y, each = n)
 	rect(rep(0:(n-1), n), y, rep(1:n, n), y - 0.5,
 	     col = cols[, subset], border = border)
 	palette <- palette[-subset]
 	cols    <- cols [, -subset, drop = FALSE]
     }
}

## height=5, width=12
palette.swatch()
```


## Illustration

Power of color often overestimated.
Color vision deficiencies ($\sim 8$% of male and $\sim 0.5$% of female viewers).
Other physical or technical limitations.

Time series line plot using base graphics, see Figure \ref{fig:tsplot}. Essentially:

```{r tsplot-pseudocode, eval=FALSE}
p <- c(1:3, 5)
plot(EuStockMarkets, log = "y", plot.type = "single", col = p, ...)
legend("topleft", colnames(EuStockMarkets), col = p, ...)
```

Figure also uses @color:Okabe+Ito:2008 palette and emulates a kind of red-green
color blindness known as protanopia using the \pkg{colorspace} package based on
the algorithms of @color:Machado+Oliveira+Fernandes:2009.

```{r tsplot, echo=FALSE, fig.height=9, fig.width=10, out.width='\\textwidth', fig.align='center', fig.pos='htbp', fig.cap='Time series line plot of \\code{EuStockMarkets}. Rows: Old \\code{"R3"} default palette (top), new \\code{"R4"} default palette (middle), \\code{"OkabeIto"} palette (bottom), designed to be robust under color vision deficiencies. Columns: Normal vision (left) and emulated protanope vision (right). A color legend is used in the first row and direct labels in the other rows.'}
tsplot <- function(palette = "R3", protan = FALSE, legend = TRUE, ...) {
  ## color palette
  palette <- match.arg(palette, c("R3", "R4", "Okabe-Ito"))
  p <- switch(palette,
    "R3" = palette.colors(8, "R3")[c(1:3, 5)],
    "R4" = palette.colors(8, "R4")[c(1:3, 5)],
    "Okabe-Ito" = palette.colors(4, "Okabe-Ito")[c(1, 4, 2, 3)]
  )
  
  ## emulate protanope vision?
  if(protan) p <- colorspace::protan(p) 
  
  ## time series plot
  eu <- window(EuStockMarkets, end = 1998)
  plot(eu, log = "y", ylab = "EU Stock Prices", plot.type = "single", col = p, xlim = c(1991.6, 1998.4), lwd = 1.5, ...)
  text(1991.3, 6000, palette, pos = 4, cex = 1.5)

  ## legend or direct labeling
  if(legend) {
    legend(1991.3, 6000, colnames(eu), col = p, lty = 1, lwd = 3, bty = "n")
  } else {
    text(1998, tail(eu, 1), colnames(eu), pos = 4, col = p, xpd = TRUE)
  }
  
  invisible(p)
}

par(mfrow = c(3, 2), mar = c(5, 4, 2, 1))
tsplot(palette = "R3", protan = FALSE, legend = TRUE, main = "Normal")
tsplot(palette = "R3", protan = TRUE, legend = TRUE, main = "Protanope")
par(mar = c(5, 4, 1, 1))
tsplot(palette = "R4", protan = FALSE, legend = FALSE)
tsplot(palette = "R4", protan = TRUE, legend = FALSE)
tsplot(palette = "Okabe-Ito", protan = FALSE, legend = FALSE)
tsplot(palette = "Okabe-Ito", protan = TRUE, legend = FALSE)
```

Old \code{R3} default palette: Too flashy, cyan too light, hard to distinguish for protanope viewers.

New \code{R4} default palette: Similar hues, more balanced brightness, avoid garish colors, distinguishable for viewers with color vision deficiencies.

Okabe-Ito: Designed to be robust against color vision deficiencies. Even easier to distinguish.


# HCL-based color palettes

## Construction principles

## Available palettes

```{r hcls, echo=FALSE, fig.height=9, fig.width=10, out.width='\\textwidth', fig.align='center', fig.pos='htbp', fig.cap='Sequential palettes from \\code{hcl.colors()}.'}
hcl.swatch <- function(palette = NULL, type = NULL, n = 5, nrow = 11,
  border = if (n < 15) "black" else NA, ...) {
    if(is.null(palette)) palette <- hcl.pals(type)
    cols <- sapply(palette, hcl.colors, n = n)
    ncol <- ncol(cols)
    nswatch <- min(ncol, nrow)

    par(mar = rep(0.1, 4),
        mfrow = c(1, min(6, ceiling(ncol/nrow))),
        cex = 0.7, ...)

    while (length(palette)) {
        subset <- 1:min(nrow, ncol(cols))
        plot.new()
        plot.window(c(0, n), c(0, nrow + 1))
        text(0, rev(subset) + 0.1, palette[subset], adj = c(0, 0))
        y <- rep(subset, each = n)
        rect(rep(0:(n-1), n), rev(y), rep(1:n, n), rev(y) - 0.5,
             col = cols[, subset], border = border)
        palette <- palette[-subset]
        cols <- cols[, -subset, drop = FALSE]
    }

    par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1), cex = 1)
}

hcl.swatch(type = "sequential", nrow = 12)
```

```{r hcld, echo=FALSE, fig.height=4.5, fig.width=10, out.width='\\textwidth', fig.align='center', fig.pos='htbp', fig.cap='Diverging palettes from \\code{hcl.colors()}.'}
hcl.swatch(c(hcl.pals("diverging"), hcl.pals("divergingx")), nrow = 6)
```

## Illustration

Sharpiegate.

# Concluding remarks

Mention gray shading in `hist()` and `boxplot()`?


# Comments about the style and template

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).

`rticles::rjournal_article` will help you build the correct files requirements: 

A tex file will be generated from this Rmd file and correctly included in
`RJwapper.tex` as expected to build `RJwrapper.pdf`.
All figure files will be kept in the default rmarkdown `*_files` folder. This
happens because `keep_tex = TRUE` by default in `rticles::rjournal_article`
